Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
more work in section 8 - defining sustainability

The question of what was meant by ``sustainability'' in the context of software
came up in many different parts of the workshop, specifically in one of the
keynotes, the Developing and Maintaining Software panel, and the Policy panel. \todo{check closing discussion notes for this topic}

@@

Another topic that came up during the Developing and Maintaining Software
panel is the relationship of sustainability to other software attributes.  Attendees
asked: what is the relationship between sustainability and provenance? And:
is usability separate from sustainability or a fundamental part of it?

\note{rest of section will be turned from bullets into text...}

In the Policy panel,

\begin{itemize}
\item Chris Lenhardt's presentation of Venters
  paper~\cite{Venters_WSSSPE} included:

\begin{itemize}
\item What does software sustainability mean
It ``is a rather ambiguous concept...''
\item Lack of an accepted definition gets in the way of integrating
  the concept into software engineering
\item Sustainability as a non-functional requirement
\item Quality of software architectures determines sustainability
\item Propose sustainability is a measure of central quality
  attributes: Extensibility, Interoperability, Maintainability,
  Portability, Reusability, Scalability
\item Develop an architecture evaluation framework based on scenarios
  which help to illuminate how to measure quality/sustainability at
  the architectural level
\end{itemize}

\item Chris Lenhardt's presentation of Katz paper~\cite{Katz_WSSSPE}
  included:
\begin{itemize}
\item Clarifying the notion of sustainability
\begin{itemize}
\item Provide same functionality going forward even if environment changes
\item Is functionality and usability clear
\item Will functionality be correct going into the future even if
  environment changes
\item Is the functionality desired by current and future users
\item Can it incorporate new things as they develop
\end{itemize}
\end{itemize}

\item see slide 9 of
  \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table

\item also see discussion at \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}

>>>>

The question of what was meant by ``sustainability'' in the context of software
came up in many different parts of the workshop, specifically in one of the
keynotes, the Developing and Maintaining Software panel, and the Policy panel.

@@

Another topic that came up during the Developing and Maintaining Software
panel is the relationship of sustainability to other software attributes.  Attendees
asked ``what is the relationship between sustainability and provenance?'' And,
``is usability separate from sustainability or a fundamental part of it?''

In addition, the Policy panel had a large amount of discussion about
defining sustainability, as one of the subtopics in that panel was
Modeling Sustainability, and modeling requires defining what will be modeled.

In this subtopic, two papers included discussion of the definition of
sustainability.  Venters et al.~\cite{Venters_WSSSPE} mentioned
that this is a rather ambiguous concept, and that the lack of an accepted
definition gets in the way of integrating the concept into software engineering.
They suggest that sustainability is a non-functional requirement, and that the
quality of software architectures determines sustainability.  They then
propose that sustainability is a measure of a set of central quality attributes:
extensibility, interoperability, maintainability, portability, reusability, and scalability.
Finally, they develop an architecture evaluation framework based on scenarios
that helps to illuminate how to measure quality/sustainability at
the architectural level.

Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could
be used to measure software sustainability, and though these questions might falsely lead
to yes or no answers, the complete set would determine a range of values for
sustainability. These question are:
Will the software provide same functionality going forward even if environment changes?
Are the functionality and usability clear?
Will functionality be correct going into the future even if environment changes?
Is the functionality that which is desired by current and future users?
Can the software incorporate new things as they develop?
\todo{check paper for wording - might be better there than in this version, which comes from bullets}

\todo{See slide 9 of \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table - incorporate/include table?}

\todo{Also see discussion at:} \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}
